|

### 2019-08-14 article/article.Rmd
add discussion, results

---
title:  |
        | Bottom-Up Accountability and Infrastructure Delivery: Mixed Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Fábio Barros, Guilherme Jardim Duarte, Stephen Herzog and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research assistance. This research was pre-registered on the EGAP pre-registry tool (<https://egap.org/registration/4505>) and approved by the Brazilian Ministry of Education. The authors thank the 2016 Google Social Impact Challenge for financial support and declare there are no conflict of interest. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "We study the effect of a mobile phone application designed to increase citizens' ability to monitor public works in Brazil. The app allows individuals to find delayed school constructions in 1030 Brazilian cities and send anonymous messages to politicians requesting information about construction delays and completion times. Our results show the app has a null impact on the completion times subsequently reported by the municipalities. Additionally, we find that few politicians react to citizens' requests. These findings suggest it is difficult to motivate bottom-up accountability in new democracies, especially when politicians are unresponsive to non-electoral pressures. This paper has implications for the design of bottom-up accountability mechanisms."
abstractspacing: double
keywords: accountability, Brazil, field experiment, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

We use the TDP app to conduct two experimental interventions and test its impact on six outcomes related to school completion rates and complaints to public authorities. Overall, our results show that providing information to citizens has very limited impact on policy outcomes. We find that the TDP app increased the likelihood of school completion by 6.8 percent in the first experiment, but the effect is null in the second intervention. We also find that pressure over the bureaucracy increased by 157.5 percent in our second experiment, although not in the first. Importantly, our two interventions have no significant impact on any of the four measures of school investment and completion rates we employed in this study.

The results raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. Although studies like @bjorkman2009power, @fiala2017social and @reinikka2005fighting report better policy outcomes after providing information to local communities, our findings are closer to @banerjee2010pitfalls, @keefer2014mass, @olken2007monitoring and @raffler2018weakness and offer mixed evidence that information interventions lead to greater government responsiveness. Our analysis suggests that grassroots monitoring does not affect government behaviour in Brazil, and other models of political accountability may be more effective for public service provision.

@@

The _Tá de Pé_[^tdp] (TDP) cell phone application is an initiative carried out by Transparência Brasil to foster bottom-up accountability in the Brazilian public sector. More specifically, the main goal of the TDP project is to improve responsiveness in government education expenditures. The TDP app incentivises citizens to provide up-to-date information about unfinished schools in their neighbourhoods, and that information will be assessed by a group of independent specialists. In the case the construction is behind schedule, TDP provides a writing platform whereby citizens can contact their representatives quickly and anonymously. The app then prompts citizens to write a notification to the mayor's office, which has 15 days to reply. If they do not respond to the request, the app forwards the notification to the Brazilian federal government, making it harder to the municipality to access federal funds in the future. The motivation for the intervention is that information delivery causes individuals to closely monitor public works. This, in turn, results in better social outcomes as public agents become more responsive to community demands.

@@

We built the app in the first three months of 2017 and tested it in May 2017. The first stable Android version was deployed on Google Play on 14 August 2017. A version for iOS appeared six months later. In October 2017, we started a Facebook campaign in order to publicise the app. Facebook is the most widely used social media network in Brazil with around 72 million users in 2017 [@statista2019facebook]. The campaigns cost US$ 4,479.29 and Facebook advertisements attracted 2,028 new users to the platform in October 2017 only.

@@

The first intervention was carried out from August 2017 to February 2018 using the Android version of TDP. The unit of analysis for this experiment is the Brazilian municipality. We randomly selected 150 cities to the control group and included 1030 municipalities in the treatment group. Our control condition consisted in removing all school constructions in the chosen municipalities so that citizens were unable to add information about them.

@@

where $i$ indexes the experiment units. $Y_i$ in one of the six outcomes described above, $\alpha$ is the intercept, $\beta$ denotes the average treatment effect and $T_i$ is a binary treatment indicator. $\gamma$ is a vector of fixed effects, $X_i$ is a matrix of Brazilian states' fixed effects, $\theta$ is a vector of controls and $Z_i$ an array of controls for the case $i$. The error term is denoted by $\varepsilon_i$. We cluster the standard errors at the municipality level as mayors are responsible for school investment decisions in Brazil.

@@

Table \ref{tab:results1} summarises the main results of intervention 1. All models reported here include the five control variables described in the previous section and Brazilian states' fixed effects. We also estimated the models without control variables and fixed effects and the results are unchanged. They are available in the Supplementary Materials.

@@

\newpage
\setlength{\parindent}{0cm}
\setlength{\parskip}{5pt}

>>>>

---
title:  |
        | Bottom-Up Accountability and Infrastructure Delivery: Mixed Evidence from Brazil
author: 
- "Danilo Freire[^freire]"
- "Manoel Galdino[^galdino]"
- "Umberto Mignozzetti[^mignozzetti]"
- "Jessica Voigt[^voigt]"
date: \today
thanks: "We thank Fábio Barros, Guilherme Jardim Duarte, Stephen Herzog, Catarina Roman and David Skarbek for their helpful comments. We are specially grateful to the staff of Transparência Brasil for their excellent research assistance. This research was pre-registered on the EGAP pre-registry tool (<https://egap.org/registration/4505>) and approved by the Brazilian Ministry of Education. The authors thank the 2016 Google Social Impact Challenge for financial support and declare there are no conflict of interest. Replication data and code are available at <http://github.com/umbertomignozzetti/tdp-accountability>."
abstract: "We study the effect of a mobile phone application designed to increase citizens' ability to monitor public works in Brazil. The app allows individuals to find delayed school constructions in 1030 Brazilian cities and send anonymous messages to politicians requesting information about construction delays and completion times. Our results show the app has a null impact on the completion times subsequently reported by the municipalities. Additionally, we find that few politicians react to citizens' requests. These findings suggest it is difficult to motivate bottom-up accountability in new democracies, especially when politicians are unresponsive to non-electoral pressures. This paper has implications for the design of bottom-up accountability mechanisms."
abstractspacing: double
keywords: accountability, Brazil, field experiment, impact evaluation, state capacity, technology
fontsize: 12pt
margin: 
urlcolor: darkblue
linkcolor: Mahogany
citecolor: Mahogany
spacing: double
papersize: a4paper
bibliography: references.bib
biblio-style: apalike
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    number_sections: yes
    keep_tex: no
    toc: no
    toc_depth: 3
    template: article-template.latex
---

@@

We use the TDP app to conduct two experimental interventions and test its impact on six outcomes related to school completion rates and complaints to public authorities. Overall, our results show that providing information to citizens has very limited impact on policy outcomes. We find that the TDP app increased the likelihood of school completion by 6.8 percent in the first experiment, but the effect is null in the second intervention. We also find that pressure over the bureaucracy increased by 157.5 percent in our second experiment, although not in the first. Importantly, our two interventions have no significant impact on any of the other measures of school investment and completion rates we employed in this study.

The results raise questions about the ability of citizens to hold representatives accountable using bottom-up strategies. Although studies like @bjorkman2009power, @fiala2017social and @reinikka2005fighting report better policy outcomes after providing information to local communities, our findings are closer to @banerjee2010pitfalls, @keefer2014mass, @lieberman2014does, @bjorkman2017experimental, @olken2007monitoring and @raffler2018weakness and offer mixed evidence that information interventions lead to greater government responsiveness. Our analysis suggests that grassroots monitoring does not affect government behaviour in Brazil, and other models of political accountability may be more effective for public service provision.

@@

The _Tá de Pé_[^tdp] (TDP) cell phone application is an initiative carried out by Transparência Brasil to foster bottom-up accountability in the Brazilian public sector. More specifically, the main goal of the TDP project is to improve responsiveness in government education expenditures. The TDP app incentivises citizens to provide up-to-date information about unfinished schools in their neighbourhoods, and that information will be assessed by a group of independent specialists. In the case the construction is behind schedule, TDP provides a writing platform whereby citizens can contact their representatives quickly and anonymously. The app then prompts citizens to write a notification to the mayor's office, which has 15 days to reply. If they do not respond to the request, the app forwards the notification to the Brazilian federal government, making it harder for the municipality to access federal funds in the future. The motivation for the intervention is that information delivery causes individuals to closely monitor public works. This, in turn, results in better social outcomes as public agents become more responsive to community demands.

@@

We built the app in the first three months of 2017 and tested it in May 2017. The first stable Android version was deployed on Google Play on 14 August 2017. A version for iOS came about six months later. In October 2017, we started a Facebook campaign in order to publicise the app. Facebook is the most widely used social media network in Brazil with around 72 million users in 2017 [@statista2019facebook]. The campaigns cost US$ 4,479.29 and Facebook advertisements attracted 2,028 new users to the platform in October 2017 only.

@@

The first intervention was carried out from August 2017 to February 2018 using the Android version of TDP. The unit of analysis for this experiment is the Brazilian municipality. We randomly selected 150 to the control group and included 1030 municipalities in the treatment group. Our control condition consists in removing all information about school construction from the TDP app in the chosen municipalities, so that citizens had no incentive to use the app.

@@

where $i$ indexes the experiment units. $Y_i$ is one of the six outcomes described above, $\alpha$ is the intercept, $\beta$ denotes the average treatment effect and $T_i$ is a binary treatment indicator. $\gamma$ is a vector of fixed effects, $X_i$ is a matrix of Brazilian states' fixed effects, $\theta$ is a vector of controls and $Z_i$ an array of controls for the case $i$. The error term is denoted by $\varepsilon_i$. We cluster the standard errors at the municipality level as mayors are responsible for school investment decisions in Brazil.

@@

Table \ref{tab:results1} summarises the main results of intervention 1. Each column represents the treatment effect of the TDP app on one of the six outcomes we measured for this study. All models reported here include the five control variables described in the previous section and Brazilian states' fixed effects. We also estimated the models without control variables, without fixed effects and with nearest-neighbour matching. The results are very similar to the ones below and can be found in the Supplementary Material.

\vspace{.5cm}
**TABLE 1 ABOUT HERE: RESULTS INTERVENTION 1**
\vspace{.5cm}

We find that the intervention was sufficient to improve the school completion rate. The TDP application increases in 6.85 percent the chance that a school construction is reported as finished by the end of the intervention. This is in line with our theoretical expectations, but as we discuss below, the finding is not consistent. All other coefficients are not statistically significant at conventional levels. On the one hand, the results indicate that our placebo outcome, the percentage of the invested executed before the intervention, indeed behaves as predicted. On the other hand, we expected the four remaining outcomes to increase after the introduction of the app. The literature on bottom-up accountability argues that more information about the shortcomings of public service provision will mobilise citizens who, in turn, would pressure state agents to produce better social policies. Our results do not lend support to that hypothesis.

\vspace{.5cm}
**TABLE 2 ABOUT HERE: RESULTS INTERVENTION 2**
\vspace{.5cm}

We observe a similar pattern in the second intervention. Table \ref{tab:results2} shows that the treatment does not have a statistically significant effect on any our outcomes of interest, including the placebo. This provides evidence that our previous findings are not a statistical fluke, and questions the robustness of the effect of the TDP app on school completion. Nevertheless, in our additional tests we find that the app improved the difference between the baseline and the end-line reported investment by 157.5 percent. This is a large effect, yet it only reaches statistical significance when we employ inverse probability weights with block randomization and one-to-one nearest neighbourhood matching.

Figure \ref{fig:permtest} shows the results of our randomisation inference tests. Randomisation inference allows us to estimate the probability of the sharp null hypothesis over all possible randomisations that could have occurred under our research design [@coppock2019ri2; @gerber2012field].

@@

In this paper, we discuss whether a mobile phone application can foster grassroots mobilisation and political accountability in Brazil. Our two interventions show that the results are mixed. Although we find an effect on school construction in the first intervention and on the number of complaints in the second one, the app has no consistent impact on our outcomes of interest. The findings add further evidence to the studies that cast doubts on the relationship between bottom-up accountability and local policy performance [e.g., @banerjee2010pitfalls;  @lieberman2014does; @bjorkman2017experimental; @raffler2018weakness].

What factors, then, are driving these results? It seems unlikely that the null results derive from flaws in the research design. First, our study is well powered. About 1,000 municipalities were included in the treatment groups, thus we should be able to capture average treatment effects, even small ones, if they indeed exist. Second, balance and manipulation tests indicate that the treatment allocation was successful, so we can rule out problems in the randomisation procedures. Third, after doing a series of robustness tests --- including matching and randomised inference --- we still find no evidence of treatment effect.

A plausible explanation is that individuals were unable to differentiate the effect of political corruption from those of spending cuts. Due to the severe economic crisis in 2014--2016, the Brazilian federal government introduced discretionary spending limits that affected public investment [@mello2016fix; @rossi2016impactos]. Politicians may argue that delays in school constructions are not derived from their misuse of government funds, but from the austerity measures. If this is the case, citizens will not blame local politicians for the underprovision of public goods. Consequently, representatives can dismiss individual requests as the issue is unlikely to escalate.

Moreover, it is possible that a mobile app like TDP is simply not sufficient to promote collective action. @raffler2018weakness [29] point out several reasons why information interventions might lead to higher pressure on public agents, such as citizens' apathy towards their ability to make public agents accountable, lack of official channels to support popular demands, or that constituents have more pressing needs than monitoring representatives. Furthermore, individuals may also believe they are better-off by free-riding on other users of the TDP app, thus reaping the benefits of having faster school constructions without incurring into personal costs [@olson1965logic].

In sum, our experiments suggest that popular participation and bottom-up monitoring may not be effective to improve public service delivery. Nevertheless, the null findings are informative to researchers and policy-makers. The most important recommendation derived from this study is that interventions targeting elite groups, such as lobbyists or civil servants, may render better policy outcomes than those focused at the grassroots levels. Another core lesson is that although digital interventions are promising means to deliver information, perhaps they do not have the same impact as personal, face-to-face communication. Since many developing countries share Brazil's issues with education provision, the shortcomings we describe here serve as warnings for future interventions.

\setlength{\parindent}{0cm}
\setlength{\parskip}{5pt}
